Towards Efficient First-Order Model Counting

### 2024-04-11 doc/paper.tex
Restructure the paper (and other changes)

% Use the postscript times font!
\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\urlstyle{same}

@@

\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes}
\usetikzlibrary{fit}

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem*{remark}{Remark}
\newtheorem{fact}{Fact}
% TODO: fix styling, check if everything in the header is still needed
% TODO: make sure forest arrows match non-forest arrows

@@

\DeclareMathOperator{\Dom}{Dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Preds}{Preds}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\GDR}{GDR}

@@

\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{EQ}$}

\SetKwFunction{Crane}{Crane}
\SetKwFunction{Propagate}{Propagate}
\SetKwFunction{FindBaseCases}{FindBaseCases}

\title{TBD}

% \author{%
% First Author$^1$\and
% Second Author$^2$\and
% Third Author$^{2,3}$\and
% Fourth Author$^4$ \\
% \affiliations
% $^1$First Affiliation\\
% $^2$Second Affiliation\\
% $^3$Third Affiliation\\
% $^4$Fourth Affiliation \\
% \emails
% \{first, second\}@example.com,
% third@other.example.com,
% fourth@example.com
% }

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item mention/describe the arithmetic simplification code
    \item (in Fig. 1) Crane2 doesn't use Crane, it uses itself
    \item describe which compilation rules are currently greedy or non-greedy
    \item my agenda has a plan for proving completeness for $\mathbf{C}^{2}$
    \item Can I prove that Crane's domain recursion can always simulate
          ForcLift's domain recursion?
  \end{itemize}
}

\paragraph{Changes to the compilation process.}
\begin{itemize}
  \item We make the FCGs satisfy the following property. For every domain
        $\Delta$, for every directed path $P$ without Ref nodes, the number of
        atom counting or (generalised) domain recursion or independent partial
        grounding operations on $\Delta$ on $P$ is at most one.
  \item when checking whether recursion is possible between a source and a
        target formula, the target formula (maybe the source formula as well?)
        must not include a domain as well as its subdomain. This might be
        equivalent to the formula being `unshattered'. Such formulas could be
        seen as transitory, and allowing such recursive calls leads to FCGs that
        I cannot make sense of.
  \item Implemented an if-then-else arithmetic operation as the interpretation
        of the new domain recursion node (i.e., there is one branch for when the
        domain is empty and another branch for any other domain size).
  \item \Cref{sec:main,sec:identifying} might also be outdated: there is
        probably no need to search for base cases given the new version of
        domain recursion.
\end{itemize}

\paragraph{Proof of correctness (leave for later).}
\begin{enumerate}
  \item Define what it means for a formula to be \emph{consistent}.
        \begin{itemize}
          \item The formula (including the mapping of constants to their
                domains) does not mention a domain together with its subdomain
                (i.e., all mentioned domains are pairwise disjoint).
                Specifically, we can't have a subdomain together with its parent
                domain or a subdomain together with a constant that belongs to
                its parent domain.
          \item The target of \texttt{Ref} must also be a consistent formula.
          \item The initial formula is consistent by definition.
        \end{itemize}
  \item Aim to show that Crane works correctly on all consistent formulas.
  \item Prove that, whenever some compilation rule makes the formula $\phi$
        inconsistent, greedy compilation rules will make $\phi$ consistent again
        before $\phi$ encounters non-greedy compilation rules that depend on
        $\phi$ being consistent.
        \begin{itemize}
          \item In particular, prove that shattering and unit propagation will
                always eliminate the parent domain of the domains introduced by
                atom counting.
          \item Constraint removal only applies when it can eliminate the parent
                domain entirely.
          \item Prove this for a particular assumption about which compilation
                rules are set to be greedy. For the greedy algorithm, although
                all rules are applied in a greedy manner, we still assume that
                greedy rules will be applied before non-greedy rules.
        \end{itemize}
  \item In particular, independent partial groundings and atom counting are the
        two classical rules that require the formula to be shattered.
  \end{enumerate}

\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \node at (0, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (alg1) {\Cref{alg:main}};
    \node[draw,rounded rectangle] at (6, 0) (simplification) {Simplification};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compilation};

@@

    \node[draw,fit={(alg1) (simplification) (compilation) (crane) (alg2) (alg3)},inner ysep=7pt,yshift=5pt] {};
    \node at (1.1, 0.5) {\textsc{Crane2}};

    \draw[-Latex] (formula) -- (alg1);
    \draw[-Latex] (alg1) -- node[above] {$E$} (simplification);
    \draw[-Latex] (simplification) -- node[above] {$E'$} (compilation);
    \draw[-Latex] (compilation) -- (cpp);
    \draw[-Latex] (sizes) -- (cpp);
    \draw[-Latex] (cpp) -- (count);

    \draw[-Latex,dashed] (alg1) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (alg1) -- node[midway] {uses} (alg2);
    \draw[-Latex,dashed] (alg1) -- node[midway,right] {uses} (alg3);
  \end{tikzpicture}
  \caption{Using \textsc{Crane2} to compute the model count of formula $\phi$.\
    The formula is compiled into a set of equations $E$, which are then
    algebraically simplified and compiled to a C++ program. This program can
    then be run with different command line arguments to compute the model count
    of $\phi$ for various domain sizes.}
\end{figure*}

The evaluation of base cases is done by simplifying the clauses and then using
\textsc{Crane} to find the base cases. First, while traversing the graph to find
the equations, we store two maps: $\mathcal{F}$ (which stores the mapping from
the function names to the formulae, whose model count they represent) and
$\mathcal{D}$ (which stores the mapping from the variable names to the domains
whose sizes they represent). Then, a particular domain is selected (using the
algorithm described in previous reports), and the clauses are simplified. Then,
\textsc{Crane} is called on those clauses to evaluate the base cases. After
that, we change the function names and variable to make it consistent with the
previous domain to variable mapping, and append these base cases to the set of
equations.

\paragraph{Contributions (outdated).}
\begin{itemize}
  \item \Cref{sec:identifying}
  \item \Cref{sec:simplifying}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
  \item A new set of benchmarks.
  \item The experiments for Crane and ForcLift are (going to be) more
        comprehensive than any other experimental study performed on (W)FOMC
        algorithms.
  \item Some experiments vs FastWFOMC
  \item A proof of correctness
  \item A proof of completeness
\end{itemize}

\todo[inline]{references to my previous paper and some of the \textsc{ForcLift}
  and \textsc{FastWFOMC} papers that contain experimental results}

@@

\begin{definition}
  A \emph{base case} is a function call where all arguments are either variables
  or constants (i.e., there is no subtraction within any argument), and there is
  at least one constant.
\end{definition}

@@

\begin{table*}
  \centering
  \begin{tabular}{llllll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atomic constraints\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $s = t$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|p| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in WFOMC in terms of:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item allowed quantifiers, and
      \item atomic constructions in addition to those of the form $p(t_{1}, \dots, t_{n})$ for some predicate symbol $p$ and terms $t_{1}, \dots, t_{n}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, the latter of which can depend on the domain size,
      \item $p$ is a predicate symbol, and
      \item $s$ and $t$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

\paragraph{Three types of logics}
\begin{itemize}
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item See \cref{tbl:logics} for a detailed comparison. \FO{} is used as the
        input format for \textsc{Crane} and \textsc{ForcLift}. \Ctwo{} is used
        in the literature on \textsc{FastWFOMC} and related methods (TODO:
        citations). \UFO{} is the input format supported by a privately-obtained
        version of \textsc{FastWFOMC}. Note that the publicly available version
        does not support any cardinality constraints.
  \item In the case of \textsc{Crane} and \textsc{ForcLift}, support for a
        formula as valid input does not imply that the algorithm will be able to
        compile the formula into a circuit or graph suitable for lifted model
        counting. However, it is known that \textsc{ForcLift} compilation is
        guaranteed to succeed on any \FO{} formula without constants and with at
        most two variables (TODO: citation).
\end{itemize}

@@

For any clause $C$, let $\Preds(C)$ denote the set of predicates in $C$. For any
clause or literal $l$, let $\Vars(l)$ denote the set of variables in $l$.

@@

\section{Theoretical Results}

\begin{theorem}
  The domain recursion by Dilkas and Belle~\shortcite{DBLP:conf/kr/DilkasB23}
  can simulate the domain recursion by Van den
  Broeck~\shortcite{DBLP:conf/nips/Broeck11}. This proves that Crane can (still)
  handle \FOtwo.
\end{theorem}

\begin{figure}
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\land_{r}$,circle,draw,name=dr
    [$n_{r}$],
    [$n_{x}$]
    ]
    \draw[-Latex] (dr) to [loop right] node {$\Delta \gets \Delta \setminus \{\,x\,\}$} ();
  \end{forest}
  \begin{forest}
    for tree={sn edges}
    [$\GDR(x \in \Delta)$,ellipse,draw,
    [$\land$,circle,draw,
    [$\bigwedge_{y \in \Delta}$,ellipse,draw]
    [$\land$,circle,draw]
    ]
    ]
  \end{forest}
\end{figure}

\begin{proof}
  `A theory allows for domain recursion when the theory is shattered, the theory
  contains no independent subtheories, and there exists a root binding class.'

  Stuck on some weird cases. Need to prove that the model count of $n_{x}$ will
  never depend on $|\Delta|$.
\end{proof}

\section{Smoothing (only relevant for domain recursion v3)}

@@

@@

\section{The Main Algorithm: Finding the Definitions of the Base
  Cases}\label{sec:main}

\begin{algorithm}[t]
  \caption{Finding the definitions of the base cases}\label{alg:main}
  \KwIn{formula $\phi$}
  \KwOut{a set $E$ of equations that define $B$}
  $(E, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$E$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i}$ is a constant}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $(E', \text{\textunderscore}, \text{\textunderscore}) \gets \Crane{$\psi$}$\;
    $E \gets E \cup E'$\;
  }
\end{algorithm}

See \cref{alg:main}

\todo[inline,caption={}]{
  \begin{itemize}
    \item extend it to work with base cases that are themselves recursive
    \item what does it mean for a set of equations to define a set of base cases?
    \item for base cases, we will also use the $f(\mathbf{x})$ notation
  \end{itemize}
}

\section{Identifying a Sufficient Set of Base Cases}\label{sec:identifying}

\begin{algorithm}[t]
  \caption{\protect\FindBaseCases{$E$}}\label{alg1}
  \KwIn{set $E$ of equations}
  \KwOut{set $B$ of base cases}

  $B \gets \emptyset$\;
  \ForEach{equation $(f(\mathbf{x}) = \expr{}) \in E$}{
    \ForEach{function call $f(\mathbf{x} - \mathbf{c}) \in \expr{}$}{\label{line:functioncall}
      \ForEach{$c_{i} \in \mathbf{c}$}{
        \For{$n \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
          $B \gets B \cup \{\, f(\mathbf{x})[x_{i}/n] \,\}$\;\label{line:insert}
        }
      }
    }
  }
\end{algorithm}

We know that if, say, on the RHS of all equations, the domain size appears as
$m - c_1, m - c_2, \dots, m - c_k$, then finding $f(0, x_1, x_2, \dots)$,
$f(1, x_1, x_2, \dots)$, $\dots f(m_0, x_1, x_2, \dots)$ for every function $f$,
where $m_0 = \max(c_1, c_2, \dots c_k) - 1$ forms a sufficient set of base
cases. Hence, in order to do the same efficiently, we can take that domain for
which $m_0$ is the minimum, i.e. $\argmin(\max(c_1, c_2, \dots c_k))$. Ideally,
we should calculate the base cases by finding the base cases up to
$\max(c_{1}, c_{2}, \dots) - 1$. However, currently only empty and singleton
domains are supported.

\todo[inline,caption={}]{
  \begin{itemize}
    \item NOTE: \cref{line:functioncall} iterates over all function calls in the
          algebraic expression
    \item remove the `dependencies' data structure from the description
  \end{itemize}
}

The following steps were followed while finding the base cases:

Now, we find a domain that has only terms of the form $x-1$ appearing on the RHS
of the dependencies. The base cases are then calculated by setting this domain
size to zero. For the above example, $n$ is the selected domain, and not $m$
since there are $m-2$ and $m-3$ terms appearing in the arguments.

The algorithm is described as \cref{alg1}.

\todo[inline,caption={}]{
  add an example where \cref{alg1} runs on either
  \begin{align*}
    f_{0}(m, n) &= f_{1}(m-1, n) + f_{2}(m, n-1)\\
    f_{1}(m, n) &= f_{1}(m-1, n-1) \times f_{2}(m-2, n-1)\\
    f_{2}(m, n) &= 2 \times f_{1}(m-3, n-1)
  \end{align*}
  or
  \begin{align*}
    f(m, n) &= g(m-1, n) + f(m-2, n-1)\\
    g(m, n) &= f(m-1, n-2) + g(m-1, n-1).
  \end{align*}
}

\begin{itemize}
  \item \cref{line:lim}: the limit is 2 for the arg $(x-3)$.
  \item \cref{line:insert} for $f(\mathbf{x}) = f(y, z)$, $x_{i} = y$, and
        $n = 0$, add $f(y, z)[y/0] = f(0, z)$ to $B$.
\end{itemize}

\todo[inline]{I changed the algorithm. Update the description.}

\begin{theorem}
  Under the following assumptions, \cref{alg1} is guaranteed to return a
  sufficient set of base cases:
  \begin{itemize}
    \item there is no mutual recursion (i.e., cyclic dependencies between
          functions);
    \item each function $f$ has exactly one equation with $f$ on the LHS;
    \item in the recursive definition of function $f(x_{1},\dots,x_{n})$, the
          $i$-th argument of each call to $f$ on the RHS is of the form
          $x_{i} - c_{i}$, where:
          \begin{itemize}
            \item $c_{i} \ge 0$ for all $i$, and
            \item $c_{i} > 0$ for at least one $i$.
          \end{itemize}
  \end{itemize}
\end{theorem}

\section{Propagating Domain Size Assumptions}\label{sec:simplifying}

\subsection{Motivation: Why Tautologies Are Needed and Simply Removing The
  Clauses Does Not Work}

\subsection{The Solution}

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$}}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, domain size $n \in \{\, 0, 1 \,\}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
      \If{$\Delta \in \Doms(C)$ {\bf and} $C' \ne \emptyset$}{
        $l \gets \text{an arbitrary literal in } C'$\;
        $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;
      }% TODO: find the source code and check: 1) what we're doing with unit clauses, 2) whether we just randomly pick any literal to negate
    }
  }
  \Else{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}

\todo[inline]{it would be easy to extend the second part to an arbitrary constant}

>>>>

\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\urlstyle{same}

@@

\usetikzlibrary{arrows.meta}

\newtheorem{theorem}{Theorem}
\newtheorem{fact}{Fact}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

% TODO: 9 pages!!!

@@

\DeclareMathOperator{\Dom}{Dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\GDR}{GDR}

@@

\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{EQ}$}

\newcommand{\Cranetwo}{\textsc{Crane}~2}

\SetKwFunction{Crane}{Compile}
\SetKwFunction{Propagate}{Propagate}
\SetKwFunction{FindBaseCases}{FindBaseCases}

\title{Towards Efficient First-Order Model Counting}

% TODO: check if this follows the formatting guidelines
\author{%
Ananth K. Kidambi$^1$\and
Guramrit Singh$^1$\and
Paulius Dilkas$^2$\and
Kuldeep S. Meel$^3$ \\
\affiliations
$^1$Indian Institute of Technology Bombay, India\\
$^2$National University of Singapore, Singapore\\
$^3$University of Toronto, Canada\\
\emails
\{210051002, 210050061\}@iitb.ac.in,
paulius.dilkas@nus.edu.sg,
meel@cs.toronto.edu
}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item describe which compilation rules are currently greedy or non-greedy
    \item my agenda has a plan for proving completeness for $\mathbf{C}^{2}$
    \item Can I prove that my domain recursion can always simulate the domain
          recursion of \textsc{ForcLift}?
  \end{itemize}
}

\paragraph{Papers to cite.}
\begin{itemize}
  \item original domain recursion~\cite{DBLP:conf/nips/Broeck11}
  \item recent overview paper~\cite{DBLP:conf/ijcai/Kuzelka23}
  \item conjecturing recurrence relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}
  \item algorithms
  \begin{itemize}
    \item \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
    \item \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}
    \item \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}
    \item approximate~\cite{DBLP:conf/ijcai/BremenK20}
  \end{itemize}
  \item liftable fragments (TODO: perhaps good sources of benchmarks?)
  \begin{itemize}
    \item \Ctwo~\cite{DBLP:journals/jair/Kuzelka21}
    \item tree axioms~\cite{DBLP:journals/ai/BremenK23}
    \item linear order axioms~\cite{DBLP:conf/aaai/TothK23}
  \end{itemize}
  \item applications
  \begin{itemize}
    \item extensions to
          sampling~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}
    \item discovery of combinatorial
          sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23}
  \end{itemize}
\end{itemize}

% TODO: make sure to not call \Crane2 \textsc{Crane}

\paragraph{Changes to the compilation process.}
\begin{itemize}
  \item We make the FCGs satisfy the following property. For every domain
        $\Delta$, for every directed path $P$ without Ref nodes, the number of
        atom counting or (generalised) domain recursion or independent partial
        grounding operations on $\Delta$ on $P$ is at most one.
  \item when checking whether recursion is possible between a source and a
        target formula, the target formula (maybe the source formula as well?)
        must not include a domain as well as its subdomain. This might be
        equivalent to the formula being `unshattered'. Such formulas could be
        seen as transitory, and allowing such recursive calls leads to FCGs that
        I cannot make sense of.
\end{itemize}

\paragraph{Proof of correctness (leave for later).}
\begin{enumerate}
  \item Define what it means for a formula to be \emph{consistent}.
        \begin{itemize}
          \item The formula (including the mapping of constants to their
                domains) does not mention a domain together with its subdomain
                (i.e., all mentioned domains are pairwise disjoint).
                Specifically, we can't have a subdomain together with its parent
                domain or a subdomain together with a constant that belongs to
                its parent domain.
          \item The target of \texttt{Ref} must also be a consistent formula.
          \item The initial formula is consistent by definition.
        \end{itemize}
  \item Aim to show that \Cranetwo{} works correctly on all consistent formulas.
  \item Prove that, whenever some compilation rule makes the formula $\phi$
        inconsistent, greedy compilation rules will make $\phi$ consistent again
        before $\phi$ encounters non-greedy compilation rules that depend on
        $\phi$ being consistent.
        \begin{itemize}
          \item In particular, prove that shattering and unit propagation will
                always eliminate the parent domain of the domains introduced by
                atom counting.
          \item Constraint removal only applies when it can eliminate the parent
                domain entirely.
          \item Prove this for a particular assumption about which compilation
                rules are set to be greedy. For the greedy algorithm, although
                all rules are applied in a greedy manner, we still assume that
                greedy rules will be applied before non-greedy rules.
        \end{itemize}
  \item In particular, independent partial groundings and atom counting are the
        two classical rules that require the formula to be shattered.
  \end{enumerate}

 % TODO: Crane2 uses itself instead of Crane. Not really. By 'Crane', here I
 % mean the process of turning logical formulas into equations. Later on, I use
 % 'Crane' to refer to Crane2. Maybe rename this use of 'Crane' to something
 % else?
\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \node at (0, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (alg1) {\Cref{alg:main}};
    \node[draw,rounded rectangle] at (6, 0) (simplification) {Simplification};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compilation};

@@

    \node[draw,fit={(alg1) (simplification) (compilation) (crane) (alg2) (alg3)},inner ysep=7pt,yshift=5pt] {};
    \node at (1.1, 0.5) {\Cranetwo};

    \draw[-Latex] (formula) -- (alg1);
    \draw[-Latex] (alg1) -- node[above] {$\mathcal{E}$} (simplification);
    \draw[-Latex] (simplification) -- node[above] {$\mathcal{E}'$} (compilation);
    \draw[-Latex] (compilation) -- (cpp);
    \draw[-Latex] (sizes) -- (cpp);
    \draw[-Latex] (cpp) -- (count);

    \draw[-Latex,dashed] (alg1) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (alg1) -- node[midway] {uses} (alg2);
    \draw[-Latex,dashed] (alg1) -- node[midway,right] {uses} (alg3);
  \end{tikzpicture}
  \caption{Using \Cranetwo{} to compute the model count of formula $\phi$.\ The
    formula is compiled into a set of equations $\mathcal{E}$, which are then
    algebraically simplified and compiled to a C++ program. This program can
    then be run with different command line arguments to compute the model count
    of $\phi$ for various domain sizes.}
\end{figure*}

\paragraph{Contributions (outdated).}
\begin{itemize}
  \item \Cref{sec:identifying}
  \item \Cref{sec:simplifying}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
  \item Some experiments vs FastWFOMC
  \item A proof of completeness
\end{itemize}

@@

\begin{definition}
  A \emph{base case} is a function call where all arguments are either variables
  or constants (i.e., there is no subtraction within any argument), and there is
  at least one constant. We denote a base case in the same way as a signature,
  e.g., $f(\mathbf{x})$.
\end{definition}

@@

\begin{table*}
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atomic constraints\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $s = t$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|p| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in WFOMC in terms of:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item allowed quantifiers, and
      \item atomic constructions in addition to those of the form $p(t_{1}, \dots, t_{n})$ for some predicate symbol $p$ and terms $t_{1}, \dots, t_{n}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, the latter of which can depend on the domain size,
      \item $p$ is a predicate symbol, and
      \item $s$ and $t$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

\paragraph{Three types of logics}
\begin{itemize}
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item See \cref{tbl:logics} for a detailed comparison. \FO{} is used as the
        input format for \textsc{ForcLift} and its extensions \textsc{Crane} and
        \Cranetwo{}. \Ctwo{} is used in the literature on \textsc{FastWFOMC} and
        related methods (TODO: citations). \UFO{} is the input format supported
        by a privately-obtained version of \textsc{FastWFOMC}. Note that the
        publicly available version does not support any cardinality constraints.
  \item In the case of \textsc{ForcLift} and its extensions, support for a
        formula as valid input does not imply that the algorithm will be able to
        compile the formula into a circuit or graph suitable for lifted model
        counting. However, it is known that \textsc{ForcLift} compilation is
        guaranteed to succeed on any \FO{} formula without constants and with at
        most two variables (TODO: citation).
\end{itemize}

@@

For any clause or literal $l$, let $\Vars(l)$ denote the set of variables in
$l$.

@@

\section{Completing the Definitions of Recursive Functions}\label{sec:main}

\begin{algorithm}[t]
  \caption{The core part of \Cranetwo{} that compiles a formula...}\label{alg:main}% TODO: start here
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i}$ is a constant}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $(\mathcal{E}', \text{\textunderscore}, \text{\textunderscore}) \gets \Crane{$\psi$}$\;
    $\mathcal{E} \gets \mathcal{E} \cup \mathcal{E}'$\;
  }
\end{algorithm}

See \cref{alg:main}.

We write \Crane for a slight modification of the original \textsc{Crane} that
compiles a formula $\phi$ into:
\begin{itemize}
  \item a set $\mathcal{E}$ of equations,
  \item a map $\mathcal{F}$ from function names to formulas, and
  \item a map $\mathcal{D}$ from function names and parameter indices to
        domains.
\end{itemize}

% TODO: update: crane or crane2
The evaluation of base cases is done by simplifying the clauses and then using
\textsc{Crane} to find the base cases. A particular domain is selected, and the
clauses are simplified. Then, \textsc{Crane} is called on those clauses to
evaluate the base cases. After that, we change the function names and variable
to make it consistent with the previous domain to variable mapping, and append
these base cases to the set of equations.

\todo[inline,caption={}]{extend it to work with base cases that are themselves
  recursive}

\subsection{Identifying a Sufficient Set of Base Cases}\label{sec:identifying}

\begin{algorithm}[t]
  \caption{\protect\FindBaseCases{$\mathcal{E}$}}\label{alg1}
  \KwIn{set $\mathcal{E}$ of equations}
  \KwOut{set $\mathcal{B}$ of base cases}

  $\mathcal{B} \gets \emptyset$\;
  \ForEach{equation $(f(\mathbf{x}) = \expr{}) \in \mathcal{E}$}{
    \ForEach{function call $f(\mathbf{x} - \mathbf{c}) \in \expr{}$}{\label{line:functioncall}
      \ForEach{$c_{i} \in \mathbf{c}$}{
        \For{$n \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
          $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i}/n] \,\}$\;\label{line:insert}
        }
      }
    }
  }
\end{algorithm}

The algorithm is described as \cref{alg1}. We know that if, say, on the RHS of
all equations, the domain size appears as $m - c_1, m - c_2, \dots, m - c_k$,
then finding $f(0, x_1, x_2, \dots)$, $f(1, x_1, x_2, \dots)$,
$\dots f(m_0, x_1, x_2, \dots)$ for every function $f$, where
$m_0 = \max(c_1, c_2, \dots c_k) - 1$ forms a sufficient set of base cases.
Hence, in order to do the same efficiently, we can take that domain for which
$m_0$ is the minimum, i.e. $\argmin(\max(c_1, c_2, \dots c_k))$. Ideally, we
should calculate the base cases by finding the base cases up to
$\max(c_{1}, c_{2}, \dots) - 1$. However, currently only empty and singleton
domains are supported.

\begin{itemize}
  \item \cref{line:insert} for $f(\mathbf{x}) = f(y, z)$, $x_{i} = y$, and
        $n = 0$, add $f(y, z)[y/0] = f(0, z)$ to $\mathcal{B}$.
\end{itemize}

\begin{theorem}
  Under the following assumptions, \cref{alg1} is guaranteed to return a
  sufficient set of base cases:
  \begin{itemize}
    \item there is no mutual recursion
    \item each function $f$ has exactly one equation with $f$ on the LHS;
    \item in the recursive definition of function $f(x_{1},\dots,x_{n})$, the
          $i$-th argument of each call to $f$ on the RHS is of the form
          $x_{i} - c_{i}$, where:
          \begin{itemize}
            \item $c_{i} \ge 0$ for all $i$, and
            \item $c_{i} > 0$ for at least one $i$.
          \end{itemize}
  \end{itemize}
\end{theorem}

\subsection{Propagating Domain Size Assumptions}\label{sec:simplifying}

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$} TODO: find the source code and check: (1) what we're doing with unit clauses, (2) whether we just randomly pick any literal to negate}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, domain size $n \in \{\, 0, 1 \,\}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
      \If{$\Delta \in \Doms(C)$ {\bf and} $C' \ne \emptyset$}{
        $l \gets \text{an arbitrary literal in } C'$\;\label{line:smoothing1}
        $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;\label{line:smoothing2}
      }
    }
  }
  \Else{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}

We use \cref{alg2} to find the transformed formula corresponding to each base
case obtained using \cref{alg1} and call \textsc{Crane} on the formula to obtain
the required base cases.
% TODO: or crane 2?

\begin{itemize}
  \item Clauses where all literals have variables quantified over the empty
        domain are still removed.
  \item $C'$ is guaranteed to be non-empty.
  \item \Cref{line:smoothing1,line:smoothing2} are explained in
        \cref{sec:smoothingbase}.
\end{itemize}

\section{Smoothing}

@@

\subsection{Smoothing for Base Cases}\label{sec:smoothingbase}

\paragraph{Key point.}
If there is a predicate $p$ that has nothing to do with domain $\Delta$, make
sure that $p$ remains as part of the formula even if all clauses with $p$ can be
removed. Which literal is picked on \cref{line:smoothing1} of \cref{alg2} is not
important because any choice achieves the same goal: constructing a clause that
mentions the same predicates as $C'$ and is satisfied by any interpretation.

@@
